---
title: "Global Health Inequality: Descriptive Analysis"
author: "Ava Hu"
date: "February 2026"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This analysis explores global health and economic indicators sourced from the **World Health Organization's Global Health Observatory**. The dataset covers 222 countries with the following variables:

| Variable     | Description                                      |
|-------------|--------------------------------------------------|
| `country`    | Country name                                     |
| `code`       | Country code                                     |
| `gni`        | Gross National Income per capita (USD)            |
| `skulspend`  | Education expenditures per capita                 |
| `death`      | Death rate per 1,000                              |
| `healthexp`  | Health expenditures per capita                    |
| `unemp`      | Unemployment rate                                 |
| `water`      | % population with access to clean water           |
| `skul`       | % enrolled in secondary school                    |
| `maternal`   | Maternal death rate per 100,000 live births        |


## Setup

```{r library}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(psych)
library(janitor)
```

## Load Data

```{r load-data}
ind <- read_csv('data/globalind.csv', col_names = TRUE)
kable(head(ind, 10), caption = "First 10 rows of the Global Health Indicators dataset")
```

```{r structure}
str(ind)
```

The dataset contains **222 observations** across **10 variables** — 2 character fields (country, code) and 8 numeric indicators.


## Data Exploration

### Filtering by Income Level

Countries with GNI per capita exceeding $50,000, sorted by education spending:

```{r high-income}
ind %>%
  select(country, code, gni, skulspend) %>%
  filter(gni >= 50000) %>%
  arrange(desc(skulspend)) %>%
  kable(caption = "High-income countries ranked by education spending")
```

### Key Questions

**Which country has the highest health expenditure per capita?**

```{r highest-healthexp}
ind %>%
  select(country, healthexp) %>%
  arrange(desc(healthexp)) %>%
  head(10) %>%
  kable(caption = "Top 10 countries by health expenditure per capita")
```

**How many countries have unemployment rates above 10%?**

```{r high-unemployment}
high_unemp <- ind %>%
  filter(unemp > 10)

cat("Number of countries with unemployment > 10%:", nrow(high_unemp))
```

**Among countries with 100% clean water access, which has the lowest health expenditure?**

```{r water-healthexp}
ind %>%
  select(country, water, healthexp) %>%
  filter(water == 100) %>%
  arrange(healthexp) %>%
  head(10) %>%
  kable(caption = "Countries with 100% clean water access, ranked by health expenditure")
```


## Summary Statistics

```{r descriptives}
# Using the psych package for comprehensive descriptive statistics
describe(ind[3:10]) %>%
  kable(digits = 2, caption = "Descriptive statistics for all numeric indicators")
```

Notable observations:
- The **mean GNI** and **median GNI** may diverge significantly, indicating income inequality across countries.
- Missing data is present across multiple indicators — handled via `na.rm = TRUE` in calculations.


## Correlation Analysis

### Correlation Matrix

```{r correlation-matrix}
cor_matrix <- round(cor(ind[3:10], use = "pairwise.complete.obs"), 2)
kable(cor_matrix, caption = "Correlation matrix of global health indicators")
```

Key findings from the correlation matrix:
- **GNI and health expenditure** show the strongest positive correlation
- **Education spending and school enrollment** are positively correlated
- **Maternal death rate** tends to be negatively correlated with income and health spending


## Visualizations

### Distribution of Key Variables

```{r histograms, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))
hist(ind$gni, main = "Distribution of GNI per Capita",
     xlab = "GNI (USD)", col = "steelblue", border = "white")
hist(ind$healthexp, main = "Distribution of Health Expenditure",
     xlab = "Health Expenditure (USD)", col = "coral", border = "white")
```

Both distributions exhibit a **strong positive skew** — most countries cluster at lower values with a long tail toward wealthier nations. A log transformation can normalize these distributions:

```{r histograms-log, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))
hist(log10(ind$gni), main = "Log-transformed GNI",
     xlab = "log10(GNI)", col = "steelblue", border = "white")
hist(log10(ind$healthexp), main = "Log-transformed Health Expenditure",
     xlab = "log10(Health Exp)", col = "coral", border = "white")
```


### Scatter Plots

```{r scatter-gni-health, fig.width=8, fig.height=6}
plot(ind$gni, ind$healthexp,
     main = "GNI vs. Health Expenditure per Capita",
     xlab = "Gross National Income per Capita (USD)",
     ylab = "Health Expenditure per Capita (USD)",
     pch = 19,
     col = rgb(0.2, 0.4, 0.6, 0.6),
     cex = 1.2)
grid()
```

The scatter plot confirms the strong positive relationship between national income and health spending.

### Multi-variable Scatter Plot Matrix

```{r pairs-plot, fig.width=8, fig.height=8}
pairs(~ gni + healthexp + skul, data = ind,
      main = "Scatter Plot Matrix: GNI, Health Expenditure, School Enrollment")
```


## Schools Dataset Analysis

The California schools dataset provides school-level data including poverty rates, student demographics, and academic performance.

```{r load-schools}
schools <- read_csv('data/schools.csv', col_names = TRUE)
```

```{r schools-descriptives}
describe(schools[5:19]) %>%
  kable(digits = 2, caption = "Descriptive statistics for schools dataset")
```

```{r schools-correlation}
round(cor(schools[5:19], use = "pairwise.complete.obs"), 2) %>%
  kable(caption = "Correlation matrix for schools variables")
```

```{r schools-histograms, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))
hist(schools$poorpct, main = "Distribution of Poverty Rate",
     xlab = "% Students in Poverty", col = "tomato", border = "white")
hist(log10(schools$attendance), main = "Distribution of Attendance (log)",
     xlab = "log10(Attendance)", col = "mediumpurple", border = "white")
```

```{r schools-scatter, fig.width=8, fig.height=6}
plot(schools$p_moved, schools$attendanc,
     main = "Student Mobility vs. Attendance",
     xlab = "% Students Moved",
     ylab = "Attendance Rate",
     pch = 19,
     col = rgb(0.6, 0.2, 0.4, 0.5))
grid()
```

```{r schools-pairs, fig.width=8, fig.height=8}
pairs(~ poorpct + p_passed + p_moved, data = schools,
      main = "Scatter Matrix: Poverty, Pass Rate, Mobility")
```


## Key Takeaways

1. **Income drives health spending**: GNI per capita explains a large proportion of variation in health expenditure across countries.
2. **Skewed distributions**: Most global indicators are positively skewed, reflecting vast disparities between wealthy and developing nations.
3. **Education and health are linked**: Countries investing more in education tend to have better health outcomes.
4. **Poverty impacts education**: In the California schools data, poverty rates are closely tied to academic performance metrics.

---

*Data sources: WHO Global Health Observatory, California Schools Dataset*
