---
title: "Global Health Inequality: Regression Analysis"
author: "Ava Hu"
date: "February 2026"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(scipen = 999)
```

## Overview

Building on the [descriptive analysis](01_descriptive_analysis.html), this notebook explores the **predictive relationships** between global health indicators through linear regression. The primary question: *How strongly does national income predict health expenditure?*

## Setup

```{r library}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(psych)
library(janitor)
library(modelr)
```

## Load Data

```{r load-data}
ind <- read_csv('data/globalind.csv', col_names = TRUE)
schools <- read_csv('data/schools.csv', col_names = TRUE)
```


## Part I: Global Indicators

### Reviewing the Strongest Correlation

From the descriptive analysis, GNI and health expenditure showed the strongest correlation. Let's examine this relationship more closely.

```{r scatter-review, fig.width=8, fig.height=6}
plot(ind$gni, ind$healthexp,
     main = "GNI vs. Health Expenditure per Capita",
     xlab = "Gross National Income per Capita (USD)",
     ylab = "Health Expenditure per Capita (USD)",
     pch = 19,
     col = rgb(0.2, 0.4, 0.6, 0.6),
     cex = 1.2)
grid()
```


### Simple Linear Regression: GNI → Health Expenditure

A linear model with `healthexp` as the dependent variable and `gni` as the independent variable:

```{r model-1}
ind_mod1 <- lm(formula = healthexp ~ gni, data = ind)
summary(ind_mod1)
```

**Interpretation:**

- **R² = 0.87** — 87% of the variation in health expenditure is explained by GNI alone.
- **p-value < 0.05** — The relationship is statistically significant.
- **GNI coefficient ≈ 0.107** — For every $1 increase in GNI per capita, health expenditure increases by approximately $0.11. In practical terms: *for every $10 increase in GNI, health spending rises by about $1.*


### Multiple Regression: Adding Unemployment

Does unemployment improve the model?

```{r model-2}
ind_mod2 <- lm(formula = healthexp ~ gni + unemp, data = ind)
summary(ind_mod2)
```

**Result:** Adding unemployment does **not** improve the model — the adjusted R² remains at 0.87 and unemployment's p-value (0.265) is not significant. GNI alone is sufficient.


## Part II: California Schools

### Exploring Poverty and Academic Performance

```{r schools-exploration, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))
hist(schools$poorpct, main = "Distribution of Poverty Rate",
     xlab = "% Students in Poverty", col = "tomato", border = "white")
hist(schools$p_passed, main = "Distribution of Pass Rate",
     xlab = "% Students Passed", col = "mediumpurple", border = "white")
```

```{r schools-scatter, fig.width=8, fig.height=6}
plot(schools$poorpct, schools$p_passed,
     main = "Schools with Higher Poverty Rates Have Lower Pass Rates",
     xlab = "Percent of Students in Poverty",
     ylab = "Percent Passed",
     pch = 19,
     col = rgb(0.2, 0.4, 0.6, 0.6),
     cex = 1.2)
grid()
```


### Regression: Poverty → Pass Rate

```{r schools-model-1}
schools_mod1 <- lm(formula = p_passed ~ poorpct, data = schools)
summary(schools_mod1)
```

```{r schools-plot-regression, fig.width=8, fig.height=6}
plot(
  schools$poorpct,
  schools$p_passed,
  main = "Schools with Higher Poverty Rates Tend to Have Lower Pass Rates",
  xlab = "Percent of Students in Poverty",
  ylab = "Percent Passed",
  pch = 19,
  col = rgb(0.2, 0.4, 0.6, 0.6),
  cex = 1.2
)
abline(schools_mod1, col = "darkred", lwd = 2)
grid()
```


### Multiple Regression: Poverty + Minority Percentage

```{r schools-model-2}
schools_mod2 <- lm(formula = p_passed ~ poorpct + p_minority, data = schools)
summary(schools_mod2)
```


### Residual Analysis

Adding predicted values and residuals to identify outlier schools that perform significantly above or below expectations given their poverty levels:

```{r residuals}
schools_pr <- schools %>%
  add_predictions(schools_mod1) %>%
  add_residuals(schools_mod1)

# Schools performing significantly ABOVE expectations (positive residuals)
outliers_positive <- schools_pr %>%
  filter(resid > 2 * sd(resid, na.rm = TRUE)) %>%
  arrange(desc(resid))

# Schools performing significantly BELOW expectations (negative residuals)
outliers_negative <- schools_pr %>%
  filter(resid < -2 * sd(resid, na.rm = TRUE)) %>%
  arrange(resid)
```

**Schools outperforming expectations** (pass rate much higher than predicted by poverty level):

```{r show-positive}
kable(outliers_positive %>% select(Name, poorpct, p_passed, pred, resid),
      col.names = c("School", "Poverty %", "Pass Rate", "Predicted", "Residual"),
      digits = 1,
      caption = "Schools with large positive residuals")
```

**Schools underperforming expectations** (pass rate much lower than predicted):

```{r show-negative}
kable(outliers_negative %>% select(Name, poorpct, p_passed, pred, resid),
      col.names = c("School", "Poverty %", "Pass Rate", "Predicted", "Residual"),
      digits = 1,
      caption = "Schools with large negative residuals")
```

```{r residual-plot, fig.width=8, fig.height=6}
ggplot(schools_pr, aes(x = poorpct, y = resid)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkred") +
  labs(title = "Residual Plot: Poverty vs. Model Residuals",
       x = "Percent of Students in Poverty",
       y = "Residual (Actual - Predicted)") +
  theme_minimal()
```

---

*Data sources: WHO Global Health Observatory, California Schools Dataset*
