---
title: "U.S. Census Demographics: Race & Age in California"
author: "Ava Hu"
date: "February 2026"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This analysis uses the **U.S. Census Bureau's American Community Survey (ACS)** to examine demographic patterns across California — from county-level summaries to census tract-level detail. Using the `tidycensus` R package, we pull data directly from the Census API and calculate racial/ethnic composition and age distributions.

**Focus areas:**
- Median age by sex across California counties
- Racial and ethnic composition at the census tract level (Alameda County)
- County-level demographic comparisons
- Joining multi-year Census data to track changes


## Setup

```{r library}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(psych)
library(tidycensus)
```

> **Note:** This analysis requires a Census API key. You can request one at [api.census.gov](https://api.census.gov/data/key_signup.html) and install it with `census_api_key("YOUR_KEY", install = TRUE)`.


## Part I: Age Demographics

### County-Level Median Age

Pulling median age data for all California counties from the 2023 ACS 5-year estimates:

```{r ca-median-age}
ca <- get_acs(geography = "county",
              variables = "B01002_001",
              state = "CA",
              year = 2023)

ca %>%
  arrange(desc(estimate)) %>%
  head(15) %>%
  kable(caption = "California counties with the highest median age (2023 ACS)")
```


### Tract-Level Age by Sex (Alameda County)

Pulling median age disaggregated by sex at the census tract level:

```{r alameda-age}
agevars <- c("B01002_001", "B01002_002", "B01002_003")

altrtage <- get_acs(geography = "tract",
                    variables = agevars,
                    county = "Alameda",
                    state = "CA",
                    year = 2023)

# Pivot from stacked to wide format
wide_altrtage <- altrtage %>%
  pivot_wider(names_from = variable, values_from = c(estimate, moe))

# Rename columns for clarity
pretty_altrtage <- wide_altrtage %>%
  dplyr::select(GEOID, NAME,
         medage = estimate_B01002_001,
         male_age = estimate_B01002_002,
         fem_age = estimate_B01002_003)

kable(head(pretty_altrtage, 15),
      caption = "Median age by sex for Alameda County census tracts")
```


## Part II: Racial & Ethnic Composition

### Alameda County — Census Tract Level

```{r alameda-race}
racevars <- c("B03002_001", "B03002_002", "B03002_003", "B03002_004",
              "B03002_005", "B03002_006", "B03002_007", "B03002_008",
              "B03002_009", "B03002_012")

alameda <- get_acs(geography = "tract",
                   variables = racevars,
                   state = "CA",
                   county = "Alameda",
                   year = 2023)

# Pivot and rename
wide_alameda <- alameda %>%
  pivot_wider(names_from = variable, values_from = c(estimate, moe))

pretty_alameda <- wide_alameda %>%
  dplyr::select(GEOID, NAME,
         pop = estimate_B03002_001,
         not_hisp = estimate_B03002_002,
         white = estimate_B03002_003,
         black = estimate_B03002_004,
         am_ind = estimate_B03002_005,
         asian = estimate_B03002_006,
         pac_isl = estimate_B03002_007,
         other = estimate_B03002_008,
         mult = estimate_B03002_009,
         hispanic = estimate_B03002_012)
```

### Computing Demographic Percentages

```{r alameda-pct}
alameda_fin <- pretty_alameda %>%
  mutate(pct_white = round(100 * (white / pop), 1)) %>%
  mutate(pct_black = round(100 * (black / pop), 1)) %>%
  mutate(pct_am_ind = round(100 * (am_ind / pop), 1)) %>%
  mutate(pct_asian = round(100 * (asian / pop), 1)) %>%
  mutate(pct_pac_isl = round(100 * (pac_isl / pop), 1)) %>%
  mutate(pct_other = round(100 * (other / pop), 1)) %>%
  mutate(pct_mult = round(100 * (mult / pop), 1)) %>%
  mutate(pct_hisp = round(100 * (hispanic / pop), 1))

kable(head(alameda_fin %>%
             select(NAME, pop, pct_white, pct_black, pct_asian, pct_hisp, pct_mult), 15),
      caption = "Racial/ethnic composition by census tract (Alameda County)")
```


### California Counties — Statewide Comparison

```{r ca-race-counties}
cacounties <- get_acs(geography = "county",
                      variables = racevars,
                      state = "CA",
                      year = 2023)

wide_cacounties <- cacounties %>%
  pivot_wider(names_from = variable, values_from = c(estimate, moe))

pretty_cacounties <- wide_cacounties %>%
  dplyr::select(GEOID, NAME,
         pop = estimate_B03002_001,
         not_hisp = estimate_B03002_002,
         white = estimate_B03002_003,
         black = estimate_B03002_004,
         am_ind = estimate_B03002_005,
         asian = estimate_B03002_006,
         pac_isl = estimate_B03002_007,
         other = estimate_B03002_008,
         mult = estimate_B03002_009,
         hispanic = estimate_B03002_012)

cacounties_fin <- pretty_cacounties %>%
  mutate(pct_white = round(100 * (white / pop), 1)) %>%
  mutate(pct_black = round(100 * (black / pop), 1)) %>%
  mutate(pct_am_ind = round(100 * (am_ind / pop), 1)) %>%
  mutate(pct_asian = round(100 * (asian / pop), 1)) %>%
  mutate(pct_pac_isl = round(100 * (pac_isl / pop), 1)) %>%
  mutate(pct_other = round(100 * (other / pop), 1)) %>%
  mutate(pct_mult = round(100 * (mult / pop), 1)) %>%
  mutate(pct_hisp = round(100 * (hispanic / pop), 1))

kable(head(cacounties_fin %>%
             select(NAME, pop, pct_white, pct_black, pct_asian, pct_hisp) %>%
             arrange(desc(pop)), 20),
      caption = "Racial/ethnic composition by county (California, 2023)")
```


## Part III: Joining Multi-Year Census Data

Comparing household income across years to track economic changes:

```{r join-years}
ca22 <- get_acs(geography = "county",
                variables = "B19013_001",
                state = "CA",
                year = 2022)

ca23 <- get_acs(geography = "county",
                variables = "B19013_001",
                state = "CA",
                year = 2023)

# Join by GEOID and compute year-over-year change
ca_joined <- ca23 %>%
  inner_join(ca22, by = "GEOID") %>%
  select(GEOID, county = NAME.x,
         hhi_2023 = estimate.x,
         hhi_2022 = estimate.y) %>%
  mutate(change = hhi_2023 - hhi_2022,
         pct_change = round(100 * change / hhi_2022, 1)) %>%
  arrange(desc(pct_change))

kable(head(ca_joined, 15),
      caption = "Counties with the largest household income increases (2022–2023)")
```

```{r join-bottom}
kable(tail(ca_joined %>% arrange(desc(pct_change)), 10),
      caption = "Counties with the smallest or negative household income changes (2022–2023)")
```


## Key Takeaways

1. **California is highly diverse**: Racial composition varies dramatically between counties and even within counties at the tract level.
2. **Age patterns vary geographically**: Median age differs significantly across tracts, reflecting different community characteristics.
3. **Census data enables temporal analysis**: By joining multi-year ACS data, we can track economic changes at the county level over time.
4. **The `tidycensus` pipeline**: Pull → Pivot → Select/Rename → Mutate provides a repeatable workflow for Census data analysis.

---

*Data source: U.S. Census Bureau, American Community Survey 5-Year Estimates (2022–2023)*
