---
title: "J 298"
author: "Ava Hu"
date: "Feb. 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Library (get them out of your toolbox)

```{r library}
library(tidyverse)
library(dplyr) 
library(ggplot2) 
library(knitr) 
library(psych) 
library(janitor)
library(tidycensus)
```


## Load data


```
ind <- read_csv('globalind.csv', col_names = TRUE) # import global indicator data
view(ind) #view indicator data

str(ind)


```

#Select

```
ind %>% select(country,code,gni) # select three fields
ind  %>% select(country,code,gni) %>% 
              arrange(desc(gni)) #select three fields and sort by gni descending
              
              
sally <- ind %>% 
    select(country,code,gni) %>% 
    arrange(desc(gni))



```

#Filter

```
filter(ind, country == "United States") # results show up in console below 

filter(ind, gni > 60000)


sally <- ind %>% 
  filter(gni > 30000) 
  
```

#Put it all together

```
ind %>% 
  select(country, code, gni, skulspend) %>% 
  filter(gni >= 50000) %>% 
  arrange(desc(skulspend))
  
  
  
ind %>% 
  select(country,code, gni, skulspend) %>% 
  filter(gni >= 50000) %>% 
  arrange(desc(skulspend))
  
  
```
#Practice
1. Which country has the highest healthexp?

ind %>% 
  select(country,healthexp) %>% 
  arrange(desc(healthexp))
  
answer: United States    11702.
  
2. How many countries have unemployment rates higher than 10?

ind %>% 
  select(country,unemp) %>% 
  filter(unemp > 10) %>% 
  count

56


sally <- ind %>% 
  select(country,water,healthexp) %>% 
  filter(water == 100) %>% 
  arrange(healthexp)
  



Thailand 

### Descriptives

Before diving into statistics, we need to better understand our data. What does the distribution look like? Are the mean and median close? We indicate data frame and variable like this: data_frame_name$variable_name. Let's go back to the indicator data.

```
mean(ind$gni)
median(ind$gni) 
```
Uh, oh -- We're not getting what I thought we should get. That's because there is missing data in the list of indicators by country. We need to add "na.rm=TRUE" to tell R to ignore the null values.

```
mean(ind$gni, na.rm = TRUE)
median(ind$gni,  na.rm = TRUE) 

des <- describe(ind)

```


Doing those individually is a lot of work. The 'psych' package we loaded has a handy function called 'describe'. 

```
describe(ind)
```
The first 2 fields are strings, so ignore the output for those. 

Once we've had a look at the data, we can explore it to see if there are any relationships.


### Correlations

Let's run some correlations looking at the relationship between skul (secondary school enrollment) and skulspend (spending on schools). We add "use="pairwise.complete.obs" to tell R to ignore missing data between pairs:

```
cor(ind$skul,ind$skulspend, use="pairwise.complete.obs")
```

One of the most useful stats tools is the correlation matrix, where you look at every variable against every other variable. 

In this case, let's look at columns 3 through 10. 

```
cor(ind[3:10], use="pairwise.complete.obs")
```
That's a lot of digits. If you'd like to round the output, wrap it in the round command. I'm using 2 decimal places.

```
round(cor(ind[3:10], use="pairwise.complete.obs"),2)
```

The strongest correlation is between gni and healthexp.


### Visualize your data: Histograms and scatter plots

To see the distribution of your data, draw a histogram.

```
hist(ind$gni)
hist(ind$healthexp)
```

Our results have a positive skew, meaning lots of values on the low end and a long tail to the right. In some analysis you may want to transform the results to get a stronger model.

For example, taking the log10 of variables that are positively skewed can normalize them to bolster your analysis. It is completely reasonable to do this, but it makes it more challenging to explain to readers.

hist(log10(ind$gni))
hist(log10(ind$healthexp))

To see the relationship between two variables, make a scatter plot. The independent variable is the X (listed first) and the dependent variable is the Y axis (listed second).

```
plot(ind$gni, ind$healthexp)


```


If you want to look at scatterplots for multiple variables, use the pairs command:
```
pairs(~ gni + healthexp + skul, data = ind)
```

### Practice with the schools data: 

Use schools.csv 

schools <- read_csv('schools.csv', col_names = TRUE) # import schools data


1. Descriptives - to understand the data
2. Correlation matrix - to find possible relationships
3. Histograms - on two variables from your correlation matrix
4. Scatterplot(s) - on those variables to visualize the relationship


```{r}
schools <- read_csv('schools.csv', col_names = TRUE)  #Descriptives

des_school <- describe(schools[5:19]) 

col <- corr.test(schools[5:19], use="pairwise.complete.obs")  #Correlation matrix

hist(log10(schools$attendance)) #Histograms on attendance 

```

```{r}
hist(schools$poorpct) #Histograms on poor pct
```

```{r}
plot(schools$p_moved, schools$attendanc) #plots on attendance 
```
```{r}
corr.test(schools[5:19], use="pairwise.complete.obs")
```

```{r}
pairs(~poorpct + p_passed + p_moved, data = schools)

```



